Structural and Geographic Influences on Residential Parking Availability in Toronto

### 2024-11-26 paper/paper.qmd
rendered paper and created model API

The primary estimand of interest is the conditional probability of a residential property having more than one parking space, given specific predictors such as geographic location (ward) and parking type. This estimand is operationalized as a binary outcome ("Yes" for more than one space, "No" otherwise) and modeled using predictors derived from administrative data on residential parking in Toronto. The goal is to understand how these factors jointly influence the likelihood of additional parking capacity while controlling for variability in urban layouts and zoning practices across the city.

@@

For this analysis, we used the R programming language [@citeR] to explore data on residential front yard parking in Toronto. The dataset, sourced from Toronto’s Open Data Portal [@opendata], provides detailed information on parking permissions and associated attributes across various residential areas in the city. Key aspects analyzed include the type of parking (e.g., front yard, boulevard, or widened driveway), the presence of more than one parking space, and geographic differences across city wards.

A range of `R` packages was employed for efficient data manipulation, modeling, and visualization. The `tidyverse` package suite formed the backbone of data processing, enabling seamless and reproducible workflows [@tidyverse]. File path management was handled using `here`, ensuring reliable and consistent access to the dataset [@citehere]. Data cleaning tasks were performed with `janitor`, which helped identify and rectify inconsistencies in the data [@citejanitor]. Bayesian statistical modeling was conducted using `rstanarm`, allowing for robust inference and clear interpretation of results [@stan]. The presentation of model outputs was streamlined using `modelsummary`, which facilitated the generation of concise and informative summaries [@modelsummary]. Additionally, `arrow` enabled fast and memory-efficient handling of the dataset, which was crucial for analyzing this spatially diverse data [@citearrow].

@@

Define $y_i$ as an indicator of whether a property has more than one parking space, where $y_i=1$ for properties with more than one space and $y_i=0$ otherwise. The primary predictors include $x_{1 i}$, the type of parking available on the property, and $x_{2 i}$, the geographic location (ward) of the property.

The logistic regression model is specified as follows:

$$
\begin{aligned}
y_i \mid \pi_i & \sim \operatorname{Bernoulli}\left(\pi_i\right) \\
\operatorname{logit}\left(\pi_i\right) & =\alpha+\sum_j \beta_j x_{j i}
\end{aligned}
$$

where $\pi_i$ represents the probability that property $i$ has more than one parking space, $\alpha$ is the intercept, and $\beta_j$ are the coefficients for the predictors.

The priors for the coefficients and intercept are weakly informative, set as follows:

$$
\alpha \sim \operatorname{Normal}(0,10), \quad \beta_j \sim \operatorname{Normal}(0,10)
$$

@@

# Results {@sec-results}

@@

```{r}
#| echo: false
#| eval: true
#| label: tbl-modelresults
#| tbl-cap: "Explanatory models of flight time based on wing width and wing length"
#| warning: false

>>>>

The primary estimand of interest is the conditional probability of a residential property having more than one parking space, given specific predictors such as geographic location (ward) and parking type. This estimand is operationalized through a binary outcome ("Yes" for more than one space, "No" otherwise) and modeled using predictors derived from administrative data on residential parking in Toronto. The goal is to understand how these factors jointly influence the likelihood of additional parking capacity while controlling for variability in urban layouts and zoning practices across the city.

@@

For this analysis, we used the R programming language [@citeR] to explore data on residential front yard parking in Toronto. The dataset, sourced from Toronto’s Open Data Portal [@opendata; @toronto_front_yard_parking], provides detailed information on parking permissions and associated attributes across various residential areas in the city. Key aspects analyzed include the type of parking (e.g., front yard, boulevard, or widened driveway), the presence of more than one parking space, and geographic differences across city wards.

A range of `R` packages was employed for efficient data manipulation, modeling, and visualization. The `tidyverse` package suite formed the backbone of data processing, enabling seamless and reproducible workflows [@tidyverse]. File path management was handled using `here`, ensuring reliable and consistent access to the dataset [@citehere]. Data cleaning tasks were performed with `janitor`, which helped identify and rectify inconsistencies in the data [@citejanitor]. Bayesian statistical modeling was conducted using `rstanarm`, allowing for robust inference and clear interpretation of results [@stan]. The presentation of model outputs was streamlined using `modelsummary`, which facilitated the generation of concise and informative summaries [@modelsummary]. Additionally, `arrow` enabled fast and memory-efficient handling of the dataset, which was crucial for analyzing this spatially diverse data [@citearrow]. Lastly, `testthat` was used to test the validity and integrity of the simulated and analysis datasets [@testthat].

@@

Define $y_i$ as an indicator of whether a property has more than one parking space, where $y_i=1$ for properties with more than one space and $y_i=0$ otherwise. The predictors include $x_{1 i}$, the type of parking available on the property, and $x_{2 i}$, the geographic location (ward). Parking type is included as categorical variables for "Front Yard" and "Widened Driveway," with "Boulevard Parking" serving as the reference category. Wards are included as a set of dummy variables, with Etobicoke-Lakeshore as the reference group.

The model is specified as:

$$
\begin{aligned}
& y_i \mid \pi_i \sim \operatorname{Bernoulli}\left(\pi_i\right)
\end{aligned}
$$

$$
\operatorname{logit}\left(\pi_i\right)=\alpha+\beta_1 \text { FrontYard}_i+\beta_2 \text { WidenedDriveway}_i+\gamma_1 \text { EtobicokeCentre}_i+\gamma_2 \text { ParkdaleHighPark}_i+\gamma_3 \text { TrinitySpadina}_i+\gamma_4 \text { YorkCentre}_i+ \\ \gamma_5 \text { YorkSouthWeston}_i+\gamma_6 \text { YorkWest}_i
$$

Here, $\pi_i$ represents the probability that property $i$ has more than one parking space. The intercept $\alpha$ captures the baseline log-odds for properties in Etobicoke-Lakeshore with Boulevard Parking.

Coefficients $\beta_1$ and $\beta_2$ measure the effect of "Front Yard" and "Widened Driveway" parking types relative to the reference category. The coefficients $\gamma_1, \ldots, \gamma_6$ capture the effects of specific wards compared to the baseline (Etobicoke-Lakeshore), explicitly:

- $\gamma_1$ : Etobicoke Centre
- $\gamma_2$ : Parkdale-High Park
- $\gamma_3:$ Trinity-Spadina
- $\gamma_4$ : York Centre
- $\gamma_5$ :York South-Weston
- $\gamma_6$ : York West

The priors for the intercept and coefficients are weakly informative and specified as follows:

$$
\begin{aligned}
\alpha & \sim \operatorname{Normal}(0,10) \\
\beta_1, \beta_2 & \sim \operatorname{Normal}(0,10) \\
\gamma_1, \gamma_2, \gamma_3, \gamma_4, \gamma_5, \gamma_6 & \sim \operatorname{Normal}(0,10)
\end{aligned}
$$

These priors reflect minimal prior knowledge, allowing the data to influence the posterior estimates while discouraging extreme values. The model estimates how parking type and geographic location jointly predict the likelihood of a property having more than one parking space, with the logistic link function ensuring predictions are appropriately bounded between 0 and 1.

@@

# Results {#sec-results}

@@

```{r}
#| echo: false
#| eval: true
#| label: tbl-modelresults
#| tbl-cap: "Explanatory models of having more than parking space based on parking type and city ward."
#| warning: false
#| message: false
